---
title: "BSTAT 3322 - Homework H03"
author: "by Craig W. Slinkman"
date: "September 11, 2015"
output: word_document

---

This homework has two learning objectives.  The first objective is to learn to use __RMarkdown__.  The second objective is to learn to use R-package $ggplot2$.

This homework must be a __Microsoft Word__ document generated by the RStudio knit package.  It is assumed that you have watched __R Markdown__ videos on YouTube.  We will cover __R Markdown__ Tuesday.

The points for this homework are distributed as follows:

| Question   | Requirement                 | Percent |
|------------|-----------------------------|---------|
| Overall    | Professionalism             |      20 |
| R Markdown | Word document               |      10 |
| 1          | Getting data                |      10 |
| 2          | Histograms                  |      10 |
| 3          | Descriptive statistics      |      10 |
| 4          | Analyze results             |      10 |
| 5          | Comparative boxplots        |      10 |
| 6          | Scatterplot diagram         |      10 |
| 7          | Analysis of scatterplot     |      10 |
| Total      |                             |     100 |   

You will submit your word document.  Do not hide any your chunks.


# Windfarm location analysis
Windmills (Data file: **wm1** in R-package alr4 ) 
Energy can be produced from wind using windmills. Choosing a site for a wind farm, the location of the windmills, can be a multi-millionaire dollar gamble. If wind is inadequate at the site, then the energy produced over the lifetime of the wind farm can be much less than the cost of building and operation. Prediction of long-term wind speed at a candidate site can be an important component in the decision to build or not to build. Since energy produced varies as the square of the wind speed, even small errors can have serious consequences. The data in the file **wm1** provides measurements that can be used to help in the prediction process. Data were collected every 6 hours for the year 2002, except that the month of May 2002 is missing. The values $CSpd$ are the calculated wind speeds in meters per second at a candidate site for building a wind farm. These values were collected at a tower erected on the site. The values $RSpd$ and $CSpd$ are wind speeds at a reference site, which is a nearby location for which wind speeds have been recorded over a very long time period. Airports sometimes serve as reference sites, but in this case, the reference data comes from the National Center for Environmental Modeling (NCAR, 2013). The reference is about 50 km southwest of the candidate site. Both sites are in the northern part of South Dakota. The data were provided by Mark Ahlstrom and Rolf Miller of WindLogics.  

----

#1. Getting the data
The data set **wm1 **is stored in the **alr4** package.  The steps outlined below will guide you through this process. 

1. If you have not yet installed the R-package alr4 package you will need in install it. The following R-statement will install the **alr4** package for you.  

```
install.packages( "alr4")

```
Be sure you get a conformation message that the package installed.  You do not need to include this statement in your homework R Markdown document.  

2.  Now load the package **alr4** package with either the R-command `library( alr4 )` or `require( alr4)`.  

```{r}
 
require( alr4 )     # Make package functionality available.

```

Ignore the waning messages as the effects they are complaining about do not affect anything we will do on this class.

3. Load the data.  We load the data in a package using the R-function `data()`.  This statement is demonstrated below.  

```{r}

data( wm1 )     # Make the data available to R.

```

4. Now we use the R-functions `head()` and `tail()` to confirm that we have the correct dataset.  

```{r}

head( wm1 )     # Confirm data set.
tail( wm1 )

```

----------

# 2. Histograms
Draw histograms of the variables $CSpd$ and $RSpd$.  Your histograms should look like the histograms plotted in Figure 1.  Note that you need not use the same $binwidth$ and colors I used.  Use your best judgement.  If you are unsure ask your manager.

```{r,echo=FALSE,fig.width=5}
require( alr4 )
require( ggplot2 )
data( wm1 )

ggplot( wm1,
        aes(x=RSpd)) +
    geom_histogram( aes( y=..density.. ),
                    binwidth=1,
                    color="blue",
                    fill="yellow" ) +
    geom_density( ) +
    xlab( "Reference speed (meters/second)" ) +
    ylab( "Density" ) +
    ggtitle( "Wind farm location study\nReference wind speed distribution" )

ggplot( wm1,
        aes(x=CSpd)) +
    geom_histogram( aes( y=..density.. ),
                    binwidth=1,
                    color="blue",
                    fill="yellow" ) +
    geom_density( ) +
    xlab( "Candidate speed (meters/second)" ) +
    ylab( "Density" ) +
    ggtitle( "Wind farm location study\n Candidate wind speed distribution" )

```
  
$$Figure 1$$  

----

# 3 Descriptive statistics
Use the R-functions `summary()` and `sd()` to provide numerical description of both variables in the **wm1** data set.  

----

# 4 Analyze results
Provide a short descriptive summary of each variable.  This must include the appropriate measures of location, dispersion, i.e., variability, and skewness of the sample distribution. The subsection below gives the rule for making this decision.  

## 4.1 Selection
The appropriate guide for the selection of what measures to use for measures of location (centrality) and measures of dispersion (variability) are the two rules given below.

* If the shape of the distribution is __symmetric__ and there are __no outliers__ use __mean__ as the measure location and the __standard deviation__ as the measure of variability. 

* Otherwise use the __median__ as the measure of location and the __interquartile range__ as the measure of variability.
  
----

## 5 Comparative boxplots of $RSpd$ and $CSpd$
Suppose we need to use boxplots to compare the distribution of the speed of the reference site with the site.  To do this we need to modify the shape of the data.  

Wide format is where one or more variables essentially measure the same thing.  For example in the **wm1** package the variables $RSpd$ and $CSpd$ both measure wind speed.  They appear on the same data row in the table.  

```{r,echo=FALSE}
head( wm1 )

```


To do this we will use the package reshape2. Use the **Tools** menu to install the reshape2 package.

```{r}

require( reshape2 )          # Load package to change the data from
#                            # Wide form to long form.

```

We know use the R-function melt.  Use exactly as I specify below.  That is, copy and paste it into your own script.

```{r}

rswm1 <-                                 # Change from wide form to
    melt( wm1,                           # long form data.
          variable.name = "Site",
          value.name="Speed") 
head( rswm1 )                            # Display first 6 rows.         
tail( rswm1 )                            # Display last 6 rows.

```

Now produce a comparative boxplot with the $Site$ as your variable. Follow the class standards for plots.  Use the graph plotted in Figure 2 as your guide.

```{r,echo=FALSE}

require( ggplot2 )

ggplot( rswm1,
        aes( x=Site, y=Speed)) +
    geom_boxplot() +
    xlab( "Sites" ) +
    ylab( "Speed (meters/second)" ) +
    ggtitle( "candidate and reference site wind speed distributions" )

```
  
$$Figure 2$$  

----

## 6 Additional analysis
Do the comparative boxplots add any information to you previous analysis in 4?  If SO, what do the box plots add to your analysis?   

----

## 7 Scatterplot diagram 
Plot the Candidate speed versus the reference speed for the **wm1** data.  Add rug plots to the scatterplot.  Your scatter should look almost exactly like the one given Figure 3.  

```{r,echo=FALSE,fig.height=6,fig.width=6}

require( ggplot2 )
ggplot( wm1,	
        aes( x= RSpd, y=CSpd)) +
    geom_point() +
    xlab( "Refernce site wind speed (meters/Second)" ) +
    ylab( "Candidate site wind speed (meters/second)" ) +
    ggtitle( "Windfarm location analysis" ) 

```
  
  $$Figure 3$$ 

# 8  Analysis of scatterplot
Analyze the scatterplot diagram.  Address the following issues.  

1.  Does there appear to be a linear relationship between the wind speed at the candidate site and the wind speed at the reference site?  Justify your answer. 

2. Is the relationship the wind speed at the candidate site and the wind speed at the reference site a positive or negative relationship?  Justify your answer?

3. Is the relationship the wind speed at the candidate site and the wind speed at the reference site is the relationship between a weak, moderate, or a strong relationship?  Justify your answer.  







