---
title: "BSTAT 3322 - H07"
author: "by Craig W. Slinkman"
date: "November 23, 2015"
output: word_document
---

In this homework there are 10 questions.  In each question you are asked to

* Identify the response and predictor variables. 

* Use ggplot2 to draw a suitable diagram.  

* Identify the nature of the relationship between the response and the predictor variables.  

* Form the scatterplots only discuss how strong and how useful it will be to use the predictor variable to forecast the response variable. 

## 1. Fire and Theft in Chicago
The following data concerns the relationship between the number of fires in a zip code and the number of thefts in the same zip code.  Does it appear that number of thefts be 1000 units is linearly related to the number of fires?  The data is stored in the file $Chigago.csv$. 

## 2. Radiation exposure and cancer
To measure the health consequences of this contamination, an index of exposure was calculated for each of the nine Oregon counties having frontage on either the Columbia River or the Pacific Ocean. This particular index was based on several factors, including the county's stream distance from Hanford and the average distance of its population from any water frontage. As a covariate, the cancer mortality rate was determined for each of these same counties.

The data give the index of exposure and the cancer mortality rate during 1959-1964 for the nine Oregon counties affected. Higher index values represent higher levels of contamination.


Variable	|	Description  
------------|-----------------  
County	    | Name of county  
Exposure    | Index of exposure  
Mortality   | Cancer mortality per 100,000 man-years  

The data file is a text file and you will need to read it with $read.table$ function.  I show you how do this below.

```{r Q3}
Hanford <- read.table( "Hanford.txt"   )
colnames( Hanford ) <- c( "County", "Exposure", "Mortality"  )
Hanford
```


## 3. Charting and stockprice changes
An investor believes that the signs of the zodiac in relation to his birth date control the returns of the stocks he has in his porfirio.  To obtain the data for this problem load and run the scripts $ZodiacInvestments.R$.  This will create a csv file that you can read with your R-markdown chunk.

## 4. Price and ages of Mazda cars in Melpborne Austrailia
The two columns of the data are the prices and year purchased for 124 Mazda cars, as taken from the classified section of the Melbourne Age during the course of 1991. Hence the age of the car at the time can be calculated and used to model car price.

### Comments
* The data file will need to be read with a read table statement. The read,table statement assumes the data is character so we need to transform it to numeric before computing the age.  

```{r}
Mazadas <- read.table( "Mazadas.txt ")
colnames( Mazadas ) <- c( "Year", "Price" )
Mazadas$Year        <- as.numeric(  Mazadas$Year )
Mazadas$Price       <- as.numeric( Mazadas$Price )
Mazadas$age         <- 1991 - Mazadas$Year 
```

Source

Melbourne Age, various issues during 1991.

Lee, A. (1994) Data Analysis: An introduction based on R. Auckland: Department of Statistics, University of Auckland.

# 5.  Capital and Rental Values of Auckland Properties 
The data show the capital value and annual rental value of 96 domestic properties in Auckland in 1991. The aim was to explore their relationship in the hope of being able to predict capital value from rental value, thus the latter is the explanatory variable in this case.  The data file is $Rents.csv$.  

You will need to read this data with the $read.csv$ function.  You should also divide both the response and the predictor variable by 1,000 to make your plot look better.  The data file is $Rents.dat$.

### Description

The data show the capital value and annual rental value of 96 domestic properties in Auckland in 1991. The aim was to explore their relationship in the hope of being able to predict capital value from rental value, thus the latter is the explanatory variable in this case.

```{r}
rm( list=ls() )
Units <- read.csv( "Rents.csv" )
colnames( Units ) <- c( "Capital", "Rents")
Units$Capital <- as.numeric( Units$Capital )
Units$Rents   <- as.numeric( Units$Rents )
Units <- Units[,-2]
colnames( Units ) <- c( "Capital", "Rents")
head( Units )
```


### Source

Lee, A. (1994) Data Analysis: An introduction based on R. Auckland: Department of Statistics, University of Auckland. Data courtesy of Sage Consultants Ltd.
 

# 6. Atlanta Commute Times
Install the Lock5Data package if you have not already installed it. Load the package with the $requireE$ OR $library$ functions.  Then load the $ComuteAtlanta$ with the data function.  Analyze the relationship between commute distance and compute time.  The variable the government is interested is commute time.  

# 7. Homes for Sale in California 
The dataset $HomesForSaleCA$ in the $Lock5Data$ package  contains a random sample of 30 houses for sale in California. We are interested in whether we can use number of bathrooms Baths
to predict number of bedrooms Beds in houses in California.

Lock, Robin H. (2012-10-10). Statistics: Unlocking the Power of Data, First Edition (Page 550). Wiley Higher Ed. Kindle Edition. 

#8. 1.4 Old Faithful 
(Data file: $oldfaith$  in package $alr4$) The data file gives information about eruptions of Old Faithful Geyser during October 1980. Variables are the Duration in seconds of the current eruption, and the Interval, the time in minutes to the next eruption. The data were collected by volunteers and were provided by the late Roderick Hutchinson. Apart from missing data for the period from midnight to 6 a.m., this is a complete record of eruptions for that month.

Old Faithful Geyser is an important tourist attraction, with up to several thousand people watching it erupt on pleasant summer days. The park service uses data like these to obtain a prediction equation for the time to the next eruption. Draw the relevant summary graph for predicting interval from duration and summarize your results.

# 9. Predicting Windfarm Wind Speeds
(Data file: $wm1$ in R-package $alr4$) Energy can be produced from wind using windmills. Choosing a site for a wind farm, the location of the windmills, can be a multi-millionaire dollar gamble. If wind is inadequate at the site, then the energy produced over the lifetime of the wind farm can be much less than the cost of building and operation. Prediction of long-term wind speed at a candidate site can be an important component in the decision to build or not to build. Since energy produced varies as the square of the wind speed, even small errors can have serious consequences. The data in the data frame  $wm1$ provides measurements that can be used to help in the prediction process. Data were collected every 6 hours for the year 2002, except that the month of May 2002 is missing. The values $Cspd$ are the calculated wind speeds in meters per second at a candidate site for building a wind farm. These values were collected at a tower erected on the site. The values $Rspd$ are wind speeds at a reference site, which is a nearby location for which wind speeds have been recorded over a very long time period. Airports sometimes serve as reference sites, but in this case, the reference data comes from the National Center for Environmental Modeling (NCAR, 2013). The reference is about 50 km southwest of the candidate site. Both sites are in the northern part of South Dakota. The data were provided by Mark Ahlstrom and Rolf Miller of WindLogics.

Weisberg, Sanford (2013-11-25). Applied Linear Regression (Wiley Series in Probability and Statistics) (Kindle Locations 1484-1493). Wiley. Kindle Edition. 

# MLB Sabermetrics
Suppose we want determine the average number of runs that can be purchased for a million dollars.  The code below finds all players who have more than 100 at bats in the 2014 MLB season and the players salary for the 2014 season.  The R-script below uses the Lahman MLB baseball database package stored as data frames.  The data frame is stored as $csv$ file named $SalaryWin.csv$ and will be found in this project folder.

```{r}
###########################################################
# Clear the environment.                                  #
###########################################################
#
rm( list=ls())
#
#
###########################################################
# We attempt to require the package.  If the require      #
# fails we install the package and require it again       #
###########################################################
#
success <- require( Lahman )
if( !success){
   install.packages( "Lahman" )
   require( Lahman )
}

success <- require( sqldf )

if( !success){
    install.packages( 'sqldf' )
    require( sqldf )
}
#
#
###########################################################
# We obtain the team tables for the same time period.     #
###########################################################
#
data( Teams )
TeamWins <- 
    Teams[Teams$yearID > 1999, c( 1,3, 9) ]
head( TeamWins )
tail( TeamWins )
#
#
##########################################################
# Now we get the player salaries for the same time. We   #
# Group the salaries by year and team for hoining to     #
# the Teams table.                                       #
##########################################################
#
data( Salaries )
Salaries <- Salaries[ Salaries$yearID > 1999, ]
head( Salaries )
tail( Salaries )

TeamSalaries <-
    sqldf( "Select     yearID,
                       teamID,
                       sum( salary ) / 1000000. AS teamSalary
            FROM       Salaries
            GROUP BY   yearID,
                       teamID" )
head( TeamSalaries )
tail( TeamSalaries )



###########################################################
# Now we join the TeamSalaries with the TeamWins tables   #
# using a natrual join.                                   #
###########################################################
#
SalaryWins <-
    sqldf( "
           Select * 
           FROM   TeamWins W,
                  TeamSalaries S
           WHERE W.yearID = S.yearID
             AND W.teamID = S.teamID" )

head( SalaryWins )
tail( SalaryWins )
write.csv(  SalaryWins, file="SalaryWin.csv" )
rm( list=ls() )

```
  




 


Weisberg, Sanford (2013-11-25). Applied Linear Regression (Wiley Series in Probability and Statistics) (Kindle Locations 671-674). Wiley. Kindle Edition. 