---
title: "data.table package"
author: "by Craig W. Slinkman"
date: "September 3, 2015"
output: html_document
---

Within the last year that has been an upgrade to the data.frame that is more efficient than a data.frame in programming time, processing time, and storage space. This new upgrade is called the data.table.  

## Packages

A package is a grouped set of  functionality that does not come with the base-R functionaility.  To use a package we must functionality we must install the package on our version.  Then before we use we must load the package into our R-environment. 

To load the environment we use the following menu path

$$Menu --> Tools --> Install Packages$$

We then type $data.table$ in the package box and click OK.  

to use the data.frame package we must either load the package with the $library$ function or load it with the $require$ function.  I tend to use the $require$  function.

## Loading the $data.table$ package 

```{r, library}

library( data.table )

```

Or we can use the $require$ function:

```{r}
require( data.table )
```

We will now read the $Flights.csv$ table in the data folder.  We do this to save time so we do not get networks delay.  The $data.table$ package gas a very fast read function called $fread$.  We use the $fread$ as follows:  
```{r,fread}
require( data.table )

inFile <- 
    file.path( getwd(),         # Specify path to file
               "data",
               "Flights.csv")


Flights <- fread( inFile )      # Read Flight.csv file

tables()                        # Display data about tables

Flights                         # Display some content of Flights  


```

## $data.table$ subset selection

It is simpler assnd faster to do subset selction with a $data.table$ object.  

```{r,subset}
DFW <- Flights[dest=="DFW", ]
tables()
```

## Computing statistics with $data.table$
We can compute statiustics with $data.table$'s.  We first compute the total delay as before:  
```{r,statistics}
DFW[ , tot_delay := dep_delay + arr_delay ]
with( DFW, hist( tot_delay ) )
with( DFW, boxplot( tot_delay ) )
with( DFW, summary( tot_delay ) )
```

 